**OptiFit: a fast method for fitting amplicon sequences to existing OTUs**

### 2021-10-19 paper/manuscript.tex
final paper edits

\hypertarget{taxonomic-resolution-matters-for-microbiome-based-classification-of-colorectal-cancer}{%
\section{Taxonomic Resolution Matters for Microbiome-Based
Classification of Colorectal
Cancer}\label{taxonomic-resolution-matters-for-microbiome-based-classification-of-colorectal-cancer}}

alternative titles:\\
High resolution taxonomy impedes microbiome-based classification of
colorectal cancer\\
High resolution ASVs impede microbiome-based classification of
colorectal cancer\\
Mid-level taxonomic resolution is optimal for microbiome based
classification of colorectal cancer\\
Too much of a good thing: high resolution ASVs do not improve
microbiome-based classification of colorectal cancer\\
More is not always better: high resolution ASVs do not improve
microbiome-based classification of colorectal cancer

@@

Courtney R. Armour\({^1}\), Begüm D. Topçuoğlu\(^{1,2}\), Andrea
Garretto\({^1}\), Patrick D. Schloss \({^1}\)\({^*}\)

@@

\({^2}\) Exploratory Science Center, Merck \& Co., Inc., Cambridge,
Massachusetts, USA~

@@

Colorectal cancer is a common and deadly disease in the United States
accounting for over 50,000 deaths in 2020. This progressive disease is
highly preventable with early detection and treatment, but many people
do not comply with the recommended screening guidelines. The gut
microbiome has emerged as a promising target for non-invasive detection
of colorectal cancer. Most microbiome-based classification efforts
utilize taxonomic abundance data from operational taxonomic units (OTUs)
or amplicon sequence variants (ASVs). However, it is unknown which
taxonomic resolution is optimal for microbiome-based classification of
colorectal cancer. To address this question, we used a reproducible
machine learning framework to quantify classification performance of
models based on data annotated to phylum, class, order, family, genus,
OTU, and ASV levels. We found that model performance increased with
increasing taxonomic resolution, up to the family level where
performance was equal among family, genus, and OTU data before
decreasing at the ASV level. These results are a step towards optimizing
a microbiome based diagnostic tool.

@@

Despite being highly preventable, colorectal cancer remains a leading
cause of cancer related death in the United States. Low-cost,
non-invasive detection methods could greatly improve our ability to
identify and treat early stages of disease. While the microbiome has
shown promise as a diagnostic tool, there are numerous details to refine
before a diagnostic could be widely deployed. Among these are the need
for standardization of sample collection, storage, processing, and
analysis. We address one aspect of data analysis by examining which
taxonomic level provides the optimal resolution for microbiome-based
classification of colorectal cancer. This is an important step to inform
the development of microbiome-based diagnostic tools.

@@

Colorectal cancer is one of the most common cancers in men and women and
a leading cause of cancer related deaths in the United States
(\protect\hyperlink{ref-siegel2020}{1}). Early detection and treatment
are essential to increase survival rates, but for reasons such as
invasiveness and high screening costs (i.e.~colonoscopy), many people do
not comply with recommended screening guidelines
(\protect\hyperlink{ref-garcuxeda2011}{2}). This prompts a need for low
cost, non-invasive detection methods. A growing body of research points
to the gut microbiome as a promising target for non-invasive detection
of screen relevant neoplasia (SRNs) consisting of advanced adenomas and
carcinomas (\protect\hyperlink{ref-zackular2014}{3},
\protect\hyperlink{ref-zeller2014}{4}). The diagnostic potential of the
gut microbiome in detecting SRNs has been explored through machine
learning (ML) methods using abundances of operational taxonomic unit
(OTU) classifications based on amplicon sequencing of the 16S rRNA gene
(\protect\hyperlink{ref-zackular2014}{3}). Recent work has pushed for
the use of amplicon sequence variants (ASVs) to replace OTUs for
marker-gene analysis because of the improved resolution with ASVs
(\protect\hyperlink{ref-callahan2017}{5}). However, it is unclear if
OTUs are the optimal taxonomic resolution for classifying SRNs from
microbiome data or whether the additional resolution provided by ASVs is
useful for ML classification. Topçuoğlu \emph{et al}
(\protect\hyperlink{ref-topuxe7uolu2020}{6}) recently demonstrated how
to effectively apply machine learning (ML) methods to microbiome based
classification problems and developed a framework for applying ML
practices in a more reproducible way (mikropml). This analysis utilizes
the reproducible framework developed by Topçuoğlu \emph{et al} to
determine which ML method and taxonomic level produce the best
performing classifier for detecting SRNs from microbiome data.

Utilizing publicly available 16S rRNA sequence data from stool of
patients with SRNs and healthy controls, we generated taxonomic
abundance tables with mothur (\protect\hyperlink{ref-schloss2009}{7})
annotated to phylum, class, order, family, genus, OTU and ASV levels.
Using the taxonomic abundance data and the mikropml R package
(\protect\hyperlink{ref-topuxe7uolu2021}{8}), we quantified how reliably
samples could be classified as SRN or normal using five machine learning
methods including random forest, L2-regularized logistic regression,
decision tree, gradient boosted trees (XGBoost), and support vector
machine with radial basis kernel (SVM radial). Across the five machine
learning methods tested, model performance increased with increasing
taxonomic level usually peaking around genus/OTU level before dropping
off slightly with ASVs (Supplemental Figure 1). Regardless of the
taxonomic level, random forest (RF) models consistently had the largest
area under the receiver operating characteristic curve (AUROC). Within
the RF model, the highest AUROCs were observed for family (mean AUROC:
0.689), genus (mean AUROC: 0.690), and OTU (mean AUROC: 0.693) level
data with no significant difference between the three (Figure 1A,
Supplemental Figure 2). Performance with ASVs (mean AUROC: 0.676) was
significantly lower than OTUs (p \textless{} 0.01), but comparable to
family (p = 0.06) and genus (p = 0.05) levels (Figure 1A). These results
suggest that increased resolution improves model performance up to the
OTU level where further taxonomic resolution is not necessarily better
for identifying individuals with SRNs based on microbiome composition.

While comparing AUROC values between models is a useful way to assess
the overall model performance, they summarize the performance across all
thresholds and can be misleading since models with the same AUROC can
have different ROC curve shapes (\protect\hyperlink{ref-lobo2008}{9}).
Depending on the intended implementation of the model, one may want to
optimize the sensitivity over the specificity or vice versa. In this
case, the optimal model will detect as many true positives (people with
SRNs) as possible. To further compare the model performance across
taxonomic levels we compared the sensitivity of the models at a
specificity of 90\%. The highest sensitivity values were observed for
family, genus, and OTU level data (Figure 1B), consistent with the AUROC
results. Phylum, class, order, and ASV sensitivity values were all
significantly lower than family, genus, and OTU sensitivity values
(Figure 1B). This analysis further supports the observation that finer
resolution does not improve SRN detection.

One hypothesis for the observation that model performance increases from
phylum to OTU level then drops slightly at ASV level is that at higher
taxonomic levels (e.g.~phylum) there are too few taxa and too much
overlap to reliably differentiate between cases and controls. At the
level of genus or OTU there is enough data and variation but at the ASV
level, the data is too specific to individuals and does not overlap
enough. Examination of the prevalence of taxa in samples at each level
supports this idea. A majority of taxa were present in greater than 70\%
of samples at the phylum (67\% of taxa) and class (63\% of taxa) levels.
The opposite was observed at the OTU and ASV level where 50\% and 41\%
of taxa, respectively, were only present in 20\% or less of the samples
(Supplemental Figure 3). Of note, the ML pipeline includes a
pre-processing step that occurred prior to training and classifying the
ML models. For example, perfectly correlated taxa provide the same
information to build the model and thus can be collapsed. Additionally,
features with zero or near-zero variance across samples were removed.
Interestingly, despite starting with 104,106 ASVs, only 478 (0.5\%)
remained after pre-processing. At the OTU level, 705 of the 20,079 OTUs
(3.5\%) remained after preprocessing (Table 1). While the resolution
provided by ASVs is useful in certain
contexts(\protect\hyperlink{ref-eren2013}{10},
\protect\hyperlink{ref-eren2015}{11}), these results suggest that the
resolution is too fine for use in machine learning classification of
SRNs based on microbiome composition.

@@

These results demonstrate that consideration of the appropriate
taxonomic resolution for utilizing the microbiome as a predictive tool
is warranted. In general, we found that finer taxonomic resolution
(e.g.~OTU and ASV) did not add additional sensitivity to predicting SRNs
based on microbiome composition. Family, genus, and OTU level data all
performed equally. At the ASV level the fine resolution actually impedes
model performance due to the sparsity of shared taxa and led to
decreased model performance. The tendency for ASV level annotation to
split single bacterial genomes into multiple taxa
(\protect\hyperlink{ref-schloss}{12}) could also be a contributing
factor to the sparsity of shared taxa. Additionally, these results
indicate that there are not specific individual bacterial strains that
are useful to resolve SRNs, rather sets of closely related bacterial
taxa. Overall, either family, genus, or OTU level taxonomy appear to
perform equally for predicting subjects with SRNs based on the
composition of the gut microbiome. A potential benefit of utilizing
genus or family level data could be that it may allow for merging data
generated from different 16S rRNA gene regions or sequencing platforms.

@@

\textbf{\emph{Dataset.}} Raw 16S rRNA gene amplicon sequence data
isolated from human gut samples (\protect\hyperlink{ref-baxter2016}{13})
was downloaded from NCBI Sequence Read Archive (SRP062005). This dataset
contains stool samples from 490 subjects. Based on the available
metadata, samples categorized as normal, high risk normal, or adenoma
were labeled ``normal'' for this analysis and samples categorized as
advanced adenoma or carcinoma were labeled as ``screen relevant
neoplasia'' (SRN). This resulted in a total of 261 ``normal'' samples
and 229 ``SRN'' samples.

\textbf{\emph{Data processing.}} Sequence data was processed with mothur
(1.44.3) (\protect\hyperlink{ref-schloss2009}{7}) using the SILVA
reference database (v132) (\protect\hyperlink{ref-quast2013}{14}) to
produce count tables for phylum, class, order, family, genus, OTU, and
ASV following the Schloss Lab MiSeq SOP described on the mothur website
(\url{https://mothur.org/wiki/miseq_sop/}). ASV level data was also
produced using DADA2 (\protect\hyperlink{ref-callahan2016}{15}) to
ensure consistent results with a different pipeline. Data was processed
following the DADA2 pipeline, but setting pool=TRUE to infer ASVs from
the whole dataset rather than per sample. The resulting ASV table was
subsampled for consistency with the mothur data. The results were
consistent for ASVs generated with DADA2 (mean AUROC: 0.659).

\textbf{\emph{Machine Learning.}} Machine learning models were run with
the R package mikropml (v0.0.2)
(\protect\hyperlink{ref-topuxe7uolu2021}{8}) to predict the diagnosis
category (normal vs SRN) of each sample. Data was preprocessed to
normalize values (scale/center), remove values with zero or near-zero
variance, and collapse collinear features using default parameters.
Initially the models were run with default hyperparameters, but were
expanded if the peak performance was at the edge of the hyperparameter
range. Each taxonomic model taxonomic level combination (e.g.~random
forest on genus) was run with 100 different seeds. Each seed split the
data into a training (80\%) and testing (20\%) set, and output
performance of the training and testing as area under the receiver
operating curve (AUROC).

To compare performance between taxonomic levels and models, P values
were calculated as previously described
(\protect\hyperlink{ref-topuxe7uolu2020}{6}). To compare sensitivity at
90\% specificity, probabilities on the test dataset were collected for
each seed and used to calculate sensitivity for specificity values
ranging from 0 to 1 in 0.01 increments. The sensitivity at a specificity
of 90\% was pulled for each seed. The averaged ROC curves were plotted
by taking the average and standard deviation of the sensitivity at each
specificity value. An optional output from the mikropml package is the
permuted feature importance which is quantified by iteratively permuting
each feature in the model and assessing the change in model performance.
Features are presumed to be important if the performance of the model,
measured by the AUROC, decreases when that feature is permuted. Ranking
of feature importance was determined by ordering the features based on
the average change in AUROC across the 100 seeds where features with a
larger decrease in AUROC are ranked higher in importance.

@@

@@

@@

\includegraphics[width=\textwidth,height=0.85\textheight]{figure_s4.png}\\
\textbf{Supplemental Figure 4: Top 10 important taxa at each taxonomic
level.} Summary of the 10 most important taxa for the random forest
models at each taxonomic level based on the average decrease in AUROC
when the feature is permuted. Dot represents the mean decrease in AUROC
and the lines extending from the dot represent the standard deviation
from the mean.

>>>>

\hypertarget{a-goldilocks-principle-for-the-gut-microbiome-taxonomic-resolution-matters-for-microbiome-based-classification-of-colorectal-cancer}{%
\section{A Goldilocks Principle for the Gut Microbiome: Taxonomic
Resolution Matters for Microbiome-Based Classification of Colorectal
Cancer}\label{a-goldilocks-principle-for-the-gut-microbiome-taxonomic-resolution-matters-for-microbiome-based-classification-of-colorectal-cancer}}

@@

Courtney R. Armour\({^1}\), Begüm D. Topçuoğlu\(^{1,\#}\), Andrea
Garretto\({^1}\), Patrick D. Schloss \({^1}\)\({^*}\)

@@

\({^\#}\) Current Affiliation: Exploratory Science Center, Merck \& Co.,
Inc., Cambridge, Massachusetts, USA~

@@

Colorectal cancer is a common and deadly disease in the United States
accounting for over 50,000 deaths in 2020. This progressive disease is
highly preventable with early detection and treatment, but many people
do not comply with the recommended screening guidelines. The gut
microbiome has emerged as a promising target for non-invasive detection
of colorectal cancer. Most microbiome-based classification efforts
utilize taxonomic abundance data from operational taxonomic units (OTUs)
or amplicon sequence variants (ASVs) with the goal of increasing
taxonomic resolution. However, it is unknown which taxonomic resolution
is optimal for microbiome-based classification of colorectal cancer. To
address this question, we used a reproducible machine learning framework
to quantify classification performance of models based on data annotated
to phylum, class, order, family, genus, OTU, and ASV levels. We found
that model performance increased with increasing taxonomic resolution,
up to the family level where performance was equal (p \textgreater{}
0.05) among family (mean AUC: 0.689), genus (mean AUC: 0.690), and OTU
(mean AUC: 0.693) levels before decreasing at the ASV level (p
\textless{} 0.05, mean AUC: 0.676). These results demonstrate a
trade-off between taxonomic resolution and prediction performance, where
coarse taxonomic resolution (e.g.~phylum) is not distinct enough but
fine resolution (e.g.~ASV) is to individualized to accurately classify
samples. Similar to the story of Goldilocks and the three bears,
mid-range resolution (i.e.~family, genus, and OTU) is just right for
optimal prediction of colorectal cancer from microbiome data.\\
(238 words)

@@

Despite being highly preventable, colorectal cancer remains a leading
cause of cancer related death in the United States. Low-cost,
non-invasive detection methods could greatly improve our ability to
identify and treat early stages of disease. The microbiome has shown
promise as a resource for detection of colorectal cancer. Research on
the gut microbiome tends to focus on improving our ability to profile
species and strain level taxonomic resolution. However, we found that
finer resolution impedes the ability to predict colorectal cancer based
on the gut microbiome. These results highlight the need for
consideration of the appropriate taxonomic resolution for microbiome
analyses and that finer resolution is not always more informative.\\
(110 words)

@@

Colorectal cancer is one of the most common cancers in men and women and
a leading cause of cancer related deaths in the United States (1). Early
detection and treatment are essential to increase survival rates, but
for reasons such as invasiveness and high screening costs
(i.e.~colonoscopy), many people do not comply with recommended screening
guidelines (2). This prompts a need for low cost, non-invasive detection
methods. A growing body of research points to the gut microbiome as a
promising target for non-invasive detection of screen relevant neoplasia
(SRNs) consisting of advanced adenomas and carcinomas (3, 4). The
diagnostic potential of the gut microbiome in detecting SRNs has been
explored through machine learning (ML) methods using abundances of
operational taxonomic unit (OTU) classifications based on amplicon
sequencing of the 16S rRNA gene (3). Recent work has pushed for the use
of amplicon sequence variants (ASVs) to replace OTUs for marker-gene
analysis because of the improved resolution with ASVs (5). However, it
is unclear if OTUs are the optimal taxonomic resolution for classifying
SRNs from microbiome data or whether the additional resolution provided
by ASVs is useful for ML classification. Topçuoğlu \emph{et al} (6)
recently demonstrated how to effectively apply machine learning (ML)
methods to microbiome based classification problems and developed a
framework for applying ML practices in a more reproducible way. This
analysis utilizes the reproducible framework developed by Topçuoğlu
\emph{et al} to determine which ML method and taxonomic level produce
the best performing classifier for detecting SRNs from microbiome data.

Utilizing publicly available 16S rRNA sequence data from stool of
patients with SRNs and healthy controls, we generated taxonomic
abundance tables with mothur (7) annotated to phylum, class, order,
family, genus, OTU and ASV levels. Using the taxonomic abundance data
and the mikropml R package (8), we quantified how reliably samples could
be classified as SRN or normal using five machine learning methods
including random forest, L2-regularized logistic regression, decision
tree, gradient boosted trees (XGBoost), and support vector machine with
radial basis kernel (SVM radial). Across the five machine learning
methods tested, model performance increased with increasing taxonomic
level usually peaking around genus/OTU level before dropping off
slightly with ASVs (Supplemental Figure 1). Regardless of the taxonomic
level, random forest (RF) models consistently had the largest area under
the receiver operating characteristic curve (AUROC). Within the RF
model, the highest AUROCs were observed for family (mean AUROC: 0.689),
genus (mean AUROC: 0.690), and OTU (mean AUROC: 0.693) level data with
no significant difference between the three (p \textgreater{} 0.05,
Figure 1A, Supplemental Figure 2). Performance with ASVs (mean AUROC:
0.676) was significantly lower than OTUs (p \textless{} 0.01), but
comparable to family (p = 0.06) and genus (p = 0.05) levels (Figure 1A).
These results suggest that increased resolution improves model
performance up to the OTU level where further taxonomic resolution is
not necessarily better for identifying individuals with SRNs based on
microbiome composition.

While comparing AUROC values between models is a useful way to assess
the overall model performance, they summarize the performance across all
thresholds and can be misleading since models with the same AUROC can
have different ROC curve shapes (9). Depending on the intended
implementation of the model, one may want to optimize the sensitivity
over the specificity or vice versa. In this case, the optimal model will
detect as many true positives (people with SRNs) as possible. To further
compare the model performance across taxonomic levels we compared the
sensitivity of the models at a specificity of 90\%. The highest
sensitivity values were observed for family (mean sensitivity: 0.38),
genus (mean sensitivity: 0.39), and OTU (mean sensitivity: 0.37) level
data (p \textgreater{} 0.05, Figure 1B), consistent with the AUROC
results. Phylum (mean sensitivity: 0.21), class (mean sensitivity:
0.22), order (mean sensitivity: 0.28), and ASV (mean sensitivity: 0.32)
sensitivity values were all significantly lower than family, genus, and
OTU sensitivity values (p \textless{} 0.05,Figure 1B). This analysis
further supports the observation that finer resolution does not improve
SRN detection.

One hypothesis for the observation that model performance increases from
phylum to OTU level then drops at the ASV level is that at higher
taxonomic levels (e.g.~phylum) there are too few taxa and too much
overlap to reliably differentiate between cases and controls. At the
level of genus or OTU there is enough data and variation but at the ASV
level, the data is too specific to individuals and does not overlap
enough. Examination of the prevalence of taxa in samples at each level
supports this idea. A majority of taxa were present in greater than 70\%
of samples at the phylum (67\% of taxa) and class (63\% of taxa) levels.
The opposite was observed at the OTU and ASV level where 50\% and 41\%
of taxa, respectively, were only present in 20\% or less of the samples
(Supplemental Figure 3). Of note, the ML pipeline includes a
pre-processing step that occurred prior to training and classifying the
ML models. For example, perfectly correlated taxa provide the same
information to build the model and thus can be collapsed. Additionally,
features with zero or near-zero variance across samples were removed.
Interestingly, despite starting with 104,106 ASVs, only 478 (0.5\%)
remained after pre-processing. At the OTU level, 705 of the 20,079 OTUs
(3.5\%) remained after preprocessing (Table 1). While the resolution
provided by ASVs is useful in certain contexts(10, 11), these results
suggest that the resolution is too fine for use in machine learning
classification of SRNs based on microbiome composition.

@@

These results demonstrate a Goldilocks effect such that consideration of
the appropriate taxonomic resolution for utilizing the microbiome as a
predictive tool is warranted. In general, we found that finer taxonomic
resolution (e.g.~ASV) did not add additional sensitivity to predicting
SRNs based on microbiome composition. Family, genus, and OTU level data
all performed equally. At the ASV level the fine resolution actually
impeded model performance due to the sparsity of shared taxa and led to
decreased model performance. The tendency for ASV level annotation to
split single bacterial genomes into multiple taxa (12) could also be a
contributing factor to the sparsity of shared taxa. Additionally, these
results indicate that there are not specific individual bacterial
strains that are useful to resolve SRNs, rather sets of closely related
bacterial taxa. Overall, either family, genus, or OTU level taxonomy
appear to perform equally for predicting subjects with SRNs based on the
composition of the gut microbiome. A potential benefit of utilizing
genus or family level data could be that it may allow for merging data
generated from different 16S rRNA gene regions or sequencing platforms.

@@

\textbf{\emph{Dataset.}} Raw 16S rRNA gene amplicon sequence data
isolated from human gut samples (13) was downloaded from NCBI Sequence
Read Archive (SRP062005). This dataset contains stool samples from 490
subjects. Based on the available metadata, samples categorized as
normal, high risk normal, or adenoma were labeled ``normal'' for this
analysis and samples categorized as advanced adenoma or carcinoma were
labeled as ``screen relevant neoplasia'' (SRN). This resulted in a total
of 261 ``normal'' samples and 229 ``SRN'' samples.

\textbf{\emph{Data processing.}} Sequence data was processed with mothur
(1.44.3) (7) using the SILVA reference database (v132) (14) to produce
count tables for phylum, class, order, family, genus, OTU, and ASV
following the Schloss Lab MiSeq SOP described on the mothur website
(\url{https://mothur.org/wiki/miseq_sop/}). ASV level data was also
produced using DADA2 (15) to ensure consistent results with a different
pipeline. Data was processed following the DADA2 pipeline, but setting
pool=TRUE to infer ASVs from the whole dataset rather than per sample.
The resulting ASV table was subsampled for consistency with the mothur
data. The DADA2 generated ASVs performed worse than the mothur generated
ASVs (DADA2 ASV mean AUROC: 0.659, p \textless{} 0.05).

\textbf{\emph{Machine Learning.}} Machine learning models were run with
the R package mikropml (v0.0.2) (8) to predict the diagnosis category
(normal vs SRN) of each sample. Data was preprocessed to normalize
values (scale/center), remove values with zero or near-zero variance,
and collapse collinear features using default parameters. Initially the
models were run with default hyperparameters, but were expanded if the
peak performance was at the edge of the hyperparameter range. Each
taxonomic model taxonomic level combination (e.g.~random forest on
genus) was run with 100 different seeds. Each seed split the data into a
training (80\%) and testing (20\%) set, and output performance of the
training and testing as area under the receiver operating curve (AUROC).

To compare performance between taxonomic levels and models, P values
were calculated as previously described (6). To compare sensitivity at
90\% specificity, probabilities on the test dataset were collected for
each seed and used to calculate sensitivity for specificity values
ranging from 0 to 1 in 0.01 increments. The sensitivity at a specificity
of 90\% was pulled for each seed. The averaged ROC curves were plotted
by taking the average and standard deviation of the sensitivity at each
specificity value. An optional output from the mikropml package is the
permuted feature importance which is quantified by iteratively permuting
each feature in the model and assessing the change in model performance.
Features are presumed to be important if the performance of the model,
measured by the AUROC, decreases when that feature is permuted. Ranking
of feature importance was determined by ordering the features based on
the average change in AUROC across the 100 seeds where features with a
larger decrease in AUROC are ranked higher in importance.

@@

\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}

\noindent

@@

\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}

@@

\includegraphics[width=\textwidth,height=0.85\textheight]{figure_s4.png}
\textbf{Supplemental Figure 4: Top 10 important taxa at each taxonomic
level.} Summary of the 10 most important taxa for the random forest
models at each taxonomic level based on the average decrease in AUROC
when the feature is permuted. Dot represents the mean decrease in AUROC
and the lines extending from the dot represent the standard deviation
from the mean.
